Using Path-Dependent Types to Build Type Safe JavaScript Foreign Function Interfaces

### 2014-02-13 papers/icwe14/icwe14.tex
numbers
tweaks

\usepackage[utf8]{inputenc}
\usepackage[unicode=true,hidelinks]{hyperref}
\usepackage{listings}
\usepackage{lmodern}

@@

 \institute{IRISA, Universite de Rennes, France}

@@

\begin{abstract}
The popularity of statically typed programming languages compiling to JavaScript show that there exists a fringe of the programmer population interested in leveraging the benefits of static typing to write Web applications. To be of any use, these languages need to statically expose the Web browser dynamically typed native API, which seems to be a contradiction in terms. Indeed, we observe that existing statically typed languages compiling to JavaScript expose the browser API in ways that either are not type safe, or when they are, typically over constrain the programmers. This article presents new ways to encode the challenging parts of the Web browser API in static type systems such that both type safety and expressive power are preserved. Our first encoding relies on type parameters and can be implemented in most mainstream languages but drags phantom types up to the usage sites. The second encoding does not suffer from this inconvenience but requires the support of dependent types in the language.
\end{abstract}

@@

This paper reviews some common functions of the browser API, identifies the patterns that are difficult to encode in static type systems and shows new ways to encode them in such a way that both type safety and expression power are preserved. We show that type parameters are sufficient to achieve this goal and that \emph{path-dependent types} provide an even more convenient encoding of the browser API.

@@

The main challenge comes from the fact that the parameter types and return types of these function are often too general. Indeed, functions \jscode{getElementsByTagName(name)}, \jscode{getElementById(id)} and \jscode{createElement(name)} can return values of type \jscode{DivElement} or \jscode{InputElement} or any other subtype of \jscode{Element} (their least upper bound). The interface of \jscode{Element} is more general and provides less features than its subtypes. For instance, the \jscode{ImageElement} type (representing images) has a \jscode{src} property that does not exist at the \jscode{Element} level. Similarly, the \jscode{MouseEvent} type has a \jscode{button} property that does not exist at the (more general) \jscode{Event} level, used by the function \jscode{addEventListener}.

@@

We distinguished three approaches to integrate the challenging parts of the browser API into statically typed languages. This section shows that each approach favours either type safety or expression power but none provides both type safety \emph{and} the same expression power as the native browser API. We indeed  consider that an API requiring users to do type casts is not type safe, while an API making it impossible to implement a function that can readily be implemented using the native API gives less expressive power to the programmer.

The first approach, consisting in using the least upper bound of all the possible types has the same expression power as the native browser API, but is not type safe because it sometimes requires developers to explicitly downcast values to their expected specialized type.

@@

The third approach, consisting in combining two parameters into one parameter carrying all the required information, is type safe too, but reduces the expression power because it forbids developers to partially apply the function by supplying only one parameter. Consider for instance listing~\ref{lst-js-react} that defines a function \jscode{observe} partially applying the \jscode{addEventListener} function\footnote{The code of this function has been taken (and simplified) from the existing functional reactive programming libraries Rx.js~\cite{liberty2011reactive} and Bacon.js~(\href{http://baconjs.github.io/}{http://baconjs.github.io/}).}. Such a function cannot be implemented with this approach because the name of the event and the code of the listener cannot be decoupled. Thus, this one gives less expressive power than the native browser API.

@@

In summary, the current integration of the browser API by statically typed languages compiling to JavaScript is either not type safe or not as expressive as the underlying JavaScript API. Indeed, we showed that our simple \jscode{slideshow} program requires type casts if the browser API is exposed according to the standard specification. We are able to get rid of type casts on this program by using modified browser APIs, but we presented two functions that we were not able to implement using these APIs, showing that they give less expression power than the native API.

This article aims to answer the following questions: is it possible to expose the browser API in statically typed languages in a way that both reduces the need for type casts and preserves the same expression power? What typing mechanisms do we need to achieve this? Would it be convenient to be used by end developers?

@@

In summary, our encoding is type safe and gives as much expression power as the native API since it is possible to implement exactly the same functions as we are able to implement in plain JavaScript.

@@

@@

We presented two ways to encode dynamically typed browser functions in mainstream statically typed languages like Java and Scala, using type parameters or path-dependent types. Our encodings give more type safety than existing solutions, while keeping the same expression power as the native API.

>>>>

\usepackage[unicode=true,hidelinks]{hyperref}
\usepackage{listings}
\usepackage{lmodern}

@@

 \institute{IRISA, Universit\'e de Rennes, France}

@@

\begin{abstract}
The popularity of statically typed programming languages compiling to JavaScript shows that there exists a fringe of the programmer population interested in leveraging the benefits of static typing to write Web applications. To be of any use, these languages need to statically expose the Web browser dynamically typed native API, which seems to be a contradiction in terms. Indeed, we observe that existing statically typed languages compiling to JavaScript expose the browser API in ways that either are not type safe, or when they are, typically over constrain the programmers. This article presents new ways to encode the challenging parts of the Web browser API in static type systems such that both type safety and expressive power are preserved. Our first encoding relies on type parameters and can be implemented in most mainstream languages but drags phantom types up to the usage sites. The second encoding does not suffer from this inconvenience but requires the support of dependent types in the language.
\end{abstract}

@@

This paper reviews some common functions of the browser API, identifies the patterns that are difficult to encode in static type systems and shows new ways to encode them in such a way that both type safety and expressive power are preserved. We show that type parameters are sufficient to achieve this goal and that \emph{path-dependent types} provide an even more convenient encoding of the browser API.

@@

The main challenge comes from the fact that the parameter types and return types of these functions are often too general. Indeed, functions \jscode{getElementsByTagName(name)}, \jscode{getElementById(id)} and \jscode{createElement(name)} can return values of type \jscode{DivElement} or \jscode{InputElement} or any other subtype of \jscode{Element} (their least upper bound). The interface of \jscode{Element} is more general and provides less features than its subtypes. For instance, the \jscode{ImageElement} type (representing images) has a \jscode{src} property that does not exist at the \jscode{Element} level. Similarly, the \jscode{MouseEvent} type has a \jscode{button} property that does not exist at the (more general) \jscode{Event} level, used by the function \jscode{addEventListener}.

@@

We distinguished three approaches to integrate the challenging parts of the browser API into statically typed languages. This section shows that each approach favours either type safety or expressive power but none provides both type safety \emph{and} the same expressive power as the native browser API. We indeed  consider that an API requiring users to do type casts is not type safe, while an API making it impossible to implement a function that can readily be implemented using the native API gives less expressive power to the programmer.

The first approach, consisting in using the least upper bound of all the possible types has the same expressive power as the native browser API, but is not type safe because it sometimes requires developers to explicitly downcast values to their expected specialized type.

@@

The third approach, consisting in combining two parameters into one parameter carrying all the required information, is type safe too, but reduces the expressive power because it forbids developers to partially apply the function by supplying only one parameter. Consider for instance listing~\ref{lst-js-react} that defines a function \jscode{observe} partially applying the \jscode{addEventListener} function\footnote{The code of this function has been taken (and simplified) from the existing functional reactive programming libraries Rx.js~\cite{liberty2011reactive} and Bacon.js~(\href{http://baconjs.github.io/}{http://baconjs.github.io/}).}. Such a function cannot be implemented with this approach because the name of the event and the code of the listener cannot be decoupled. Thus, this one gives less expressive power than the native browser API.

@@

In summary, the current integration of the browser API by statically typed languages compiling to JavaScript is either not type safe or not as expressive as the underlying JavaScript API. Indeed, we showed that our simple \jscode{slideshow} program requires type casts if the browser API is exposed according to the standard specification. We are able to get rid of type casts on this program by using modified browser APIs, but we presented two functions that we were not able to implement using these APIs, showing that they give less expressive power than the native API.

This article aims to answer the following questions: is it possible to expose the browser API in statically typed languages in a way that both reduces the need for type casts and preserves the same expressive power? What typing mechanisms do we need to achieve this? Would it be convenient to be used by end developers?

@@

In summary, our encoding is type safe and gives as much expressive power as the native API since it is possible to implement exactly the same functions as we are able to implement in plain JavaScript.

@@

For comparison, our single function \scalacode{addEventListener} is as type safe as the about thirty functions defined in GWT for event handling while giving more expressive power to users.

@@

We presented two ways to encode dynamically typed browser functions in mainstream statically typed languages like Java and Scala, using type parameters or path-dependent types. Our encodings give more type safety than existing solutions, while keeping the same expressive power as the native API.
